\title{Distributed Graph-Based State Space Generation}%
%
\author{Stefan Blom,\sponsor{Stefan Blom is partially sponsored by the EU under grant number FP6-NEST STREP 043235 (EC-MOAN).} %
Gijs Kant\sponsor{Gijs Kant is sponsored by the VOCHS project of the NWO.} %
and Arend Rensink\sponsor{Arend Rensink is partially sponsored by the GRAIL project of the NWO.}}
%
\institute{\email{sccblom@cs.utwente.nl} \hspace{\columnsep} \email{kant@cs.utwente.nl} \hspace{\columnsep} \email{rensink@cs.utwente.nl}\\%
Department of Computer Science\\%
University of Twente, The Netherlands}
%
\abstract{
%
\LTSMIN provides a framework in which state space generation can be distributed
easily over many cores on a single compute node, as well as over multiple
compute nodes. The tool works on the basis of a vector representation of the
states; the individual cores are assigned the task of computing all
successors of states that are sent to them. In this paper we show how this
framework can be applied in the case where states are essentially graphs
interpreted up to isomorphism, such as the ones we have been studying for
\GROOVE. This involves developing a suitable vector representation for a
canonical form of those graphs. The canonical forms are computed using a third
tool called \BLISS.
We combined the three tools to form a system for distributed state space generation
based on graph grammars.

We show that the time performance of the resulting system scales well (i.e.,
close to linear) with the number of cores.
%  This is mostly due to the fact that
% the canonical form computation, which is a very time-consuming step, is
% completely distributed. On a system with four cores or more, the
% distributed implementation beats the sequential version of \GROOVE. 
We also
report surprising statistics on the memory consumption, which imply that the
vector representation used to store graphs in \LTSMIN is
more compact than the representation used in \GROOVE.
}
%
\keywords{Graph Transformation, Symmetry Reduction, State Space Generation, Distributed Computing, \textsc{groove}, \textsc{LTSmin}}
\maketitle

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End: 
